\section{Conclusions}

This project demonstrated that the ``Sorting as Planning'' problem benefits enormously from careful constraint modeling and search strategy selection.

\subsection{Key Contributions}

\begin{enumerate}
    \item \textbf{Rich Constraint Model:} We developed a MiniZinc model with 7 specialized constraints that exploit the mathematical structure of permutations:
    \begin{itemize}
        \item Parity invariant (50\% pruning of $k$ values)
        \item Dual model with channeling (maximum propagation)
        \item Optimal swap property from group theory ($O(N)$ branching)
        \item Multiple symmetry breaking constraints
        \item Fixed point propagation
        \item Global swap uniqueness
    \end{itemize}

    \item \textbf{Intelligent Meta-Solver:} The Python controller computes the theoretical lower bound via cycle decomposition, often finding solutions on the first iteration.

    \item \textbf{Comprehensive Strategy Comparison:} We systematically compared 12 search strategies across variable selection, value selection, and restart policies.
\end{enumerate}

\subsection{Lessons Learned}

\begin{itemize}
    \item \textbf{Propagation is King:} The combination of \texttt{inverse} channeling and GAC (\texttt{alldifferent :: domain}) provides the inferential power needed for large instances.

    \item \textbf{Mathematical Insight Matters:} Group theory properties (cycle decomposition, parity) allow the meta-solver to skip entire phases of search.

    \item \textbf{No Universal Best Strategy:} Different strategies excel on different instance sizes and structures. Adaptive heuristics like \texttt{dom\_w\_deg} learn from failures and may outperform static heuristics on complex instances.

    \item \textbf{Restarts Are Essential:} Randomized value selection combined with Luby restarts prevents the solver from getting trapped in barren subtrees.
\end{itemize}

\subsection{Future Work}

Potential extensions include:
\begin{itemize}
    \item Exploiting cycle structure more explicitly (elements in different cycles should not be swapped)
    \item Portfolio solving: running multiple strategies in parallel and returning the first solution
    \item Integration with learning-based approaches for strategy selection
\end{itemize}

In conclusion, Constraint Programming is an exceptionally powerful paradigm for planning problems when the underlying mathematical structure is fully exploited through sophisticated constraint design and intelligent search control.